---
title: "APPENDIX ~ CLIMATE ANALYSIS: STA141A Final Project"
date: "April 12"
output:
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
```

# Data Loading and Preliminary Analysis
```{r}
data = na.omit(read.csv("/cloud/project/climate_change.csv"))

# For Consistency in Year Based/Month Based Analyses, we remove the first 8 months 
# of the dataset (as the data randomly begins from the 5th month of 1983)

# Therefore, our dataset starts from 1984
climate <- data[-(1:8),]

#brief summary; head()/taiL()
head(climate,10)
tail(climate,10)
summary(climate)

# structure of dataframe
str(climate)

# checking for any duplicate data (there is none)
nrow(climate)==nrow(unique(climate))

```
# Exploratory Analysis 

### FIGURE 2: Testing Correlation Between Variables

```{r}
library(GGally)
library(dplyr)
library(ggplot2)

### GROUPINGS OF THESE VECTORS EXPLAINED IN THE PAPER ###
variables <- c("Temp","Year","Month","MEI","CO2","CH4","N2O","CFC.11","CFC.12", "TSI", "Aerosols")
variables_of_interest <- c("Temp","Year","Month","CO2","CH4","N2O","CFC.11","CFC.12", "Aerosols")
variables_of_interest2 <- c("Temp","Year","CO2","CH4","N2O","CFC.11","CFC.12", "Aerosols")

# General Scatter Plot Matrix
climate %>% 
  ggpairs(columns = variables_of_interest,
           upper = list(continuous = wrap('cor', size = 3)),
           lower = list(continuous = wrap('smooth',size = .1, alpha = 0.03))) +
  theme_grey() +
  theme(axis.text = element_text(size = 3)) +
  labs(title = "FIGURE 2: Scatter Plot Matrix")

# Scatter Plot Matrix Accounting for Categorical Variable
climate %>% 
  ggpairs(columns = variables_of_interest2,
           aes(color = factor(Month)),
           upper = list(continuous = wrap('cor', size = 1)),
           lower = list(combo = list(continuous = "smooth", discrete = "boxplot"),
                        size = 0.1, alpha = 0.1),
           diag = list(continuous = wrap('densityDiag', alpha = 0.2))) + 
  theme_grey() + 
  theme(axis.text = element_text(size = 1))

# Lower Half is Correlations; Diagonal Is Density Functions; Upper Half is Corr Values


pairs(climate[variables_of_interest], main = "Pairwise Scatterplot Matrix")
```

# Testing Linear and Logistic ~ the plots are not actually used, but inform decisions
```{r}
linear_climate_model = lm(Temp ~ ., data = climate)
summary(linear_climate_model)

# Exploratory Visualization of the Response Variable ~ FIGURES 2a-2d
plot(linear_climate_model)

############## PROB NEEDS FIXING ################
binomial_climate_model <- climate %>%
  mutate(HighTemp = ifelse(Temp < mean(Temp), 1, 0)) %>%
  glm(HighTemp ~ . - Temp, data = ., family = "binomial") # deal with multi-collinearity

# balancing sensitivity and specificity of response variable
summary(binomial_climate_model)
# NEED TO USE ROC CURVE OR OTHER METHOD TO DETERMINE THRESHOLDS OR 
# CLUSTERING/HIERARCHICAL METHODS INSTEAD

##### BRIEF VISUALIZATION W/ GGPLOT #####
ggplot(climate, aes(sample = Temp)) +
  geom_qq() +
  geom_qq_line(col = "red") +
  ggtitle("Q-Q Plot of Temperature") +
  theme_minimal()
```

# Testing LDA ~ Confusion Table
```{r}
library(MASS)
lda_climate_model = lda(Temp ~ ., data = climate)
summary(lda_climate_model)

# Create a confusion matrix
predictions <- predict(lda_climate_model)
confusion_matrix <- table(Actual = climate$Temp, Predicted = predictions$class)
### WE DID NOT PRINT THE OUTPUTS AS IT IS INCOLCUSIVE AND PRINTS TOO MUCH DATA
```


# Deciding on The Best Model
```{r}
library(leaps)


#### OLD CODE THAT USES ALL VARIABLES ####
predictor_names <- names(climate)[-which(names(climate) == "Temp")]
all_subsets <- regsubsets(Temp ~ ., data = climate, 
                        nvmax = length(predictor_names), method = "exhaustive")
# Get the list of all subsets
all_subsets_list <- summary(all_subsets)$which; all_subsets_list
#### ~ you would use predictor_names in the for loop instead

included_vars <- c("Year", "MEI", "CO2", "CH4", "N2O", "CFC.11", "CFC.12", "Aerosols")
formula_str <- paste("Temp ~", paste(included_vars, collapse = " + "))
formula <- as.formula(formula_str)
all_subsets <- regsubsets(formula, data = climate, 
                          nvmax = length(included_vars), method = "exhaustive")

####### WE CAN CHANGE THE METHOD OF SEARCH but since the algorithm returns the best  
# model of each size (number of parameters 2-10 or number of predictors 1-9), 
# so the results do not depend on a penalty model for model size: it doesn’t make 
# any difference whether you want to use AIC, BIC, CIC, DIC #######

# Initialize a dataframe to store model specifications, MSE, and adjR2
model_info <- data.frame(
  model = character(),
  MSE = numeric(),
  R2 = numeric(),
  adjR2 = numeric(),
  stringsAsFactors = FALSE
)

# Extract Values for Every Model Size (1-9 predictors)
for (i in 1:length(included_vars)) {
  # Extract the coefficients for the best model of size i
  model_coef <- coef(all_subsets, id = i)
  model_formula <- as.formula(paste("Temp ~", 
                    paste(names(model_coef)[-1], collapse = "+"))) #create formulas
  # Fit Models
  fit <- lm(model_formula, data = climate)
  
  # Calculate MSE
  predictions <- predict(fit, newdata = climate)
  mse <- mean((climate$Temp - predictions)^2)
  # Get R2 Values
  r2 <- summary(fit)$r.squared
  adj_r2 <- summary(fit)$adj.r.squared
  
  # Store model information
  model_info <- rbind(model_info, data.frame(
    model = deparse(model_formula),
    MSE = mse,
    R2 = r2,
    adjR2 = adj_r2,
    stringsAsFactors = FALSE
  ))
}

model_info
```

# Selection Criteria
```{r}
# Regsubsets Identifies the Best Model at each # of predictors (1-9); 
# We decide on the best model by considering the model with the lowest MSE that 
# does not overfit the data (the adjusted R2 is not significantly smaller than the R2)

filter <- model_info[model_info$adjR2 >= (0.95 * model_info$R2), ]

# Choose the model with the lowest MSE from the filtered models
best_model <- filter[which.min(filter$MSE), ]

print(best_model)
```

# Analyzing Our Model: Figure 3

```{r}
best_model_formula <- as.formula(best_model$model)


# Fit the best model
best_fit <- lm(best_model_formula, data = climate)
# FIGURE 3 (a-d)
plot(best_fit)

# Analysis ~ t-test
summary(best_fit)
qt(1-0.05/2, 300-8-1)

# F-stat overall significance of the model
summary(best_fit)$fstatistic[1]
qf(1-0.05,df1 = 8, df2 = 300-8-1)

anova_table <- anova(best_fit)
print(anova_table)

# Calculating F-val
mean_sq_model <- sum(anova_table$`Mean Sq`[1:8])
F_value <- mean_sq_model /anova_table$`Mean Sq`[9]
# F-stat compares the variability explained by the predictors (Mean Sq Model) 
# with the variability not explained by the model (Mean Sq Residuals)
F_value
```
# F-test Lack of Fit
```{r}
residuals <- residuals(best_fit)
fitted_values <- fitted(best_fit)

# Calculate lack-of-fit sum of squares
n <- length(residuals)
mean_residuals <- mean(residuals)
lack_of_fit_ss <- sum((residuals - mean_residuals)^2)

# Calculate residual sum of squares
residual_ss <- sum(residuals^2)

# Degrees of freedom for the lack-of-fit test
df_lack_of_fit <- n - length(coefficients(best_fit)) #Adjust for number of coefficients 

# Degrees of freedom for residuals
df_residuals <- df.residual(best_fit)

# Calculate F-value for lack of fit
F_lack_of_fit <- (lack_of_fit_ss / df_lack_of_fit) / (residual_ss / df_residuals)
F_critical <- qf(1 - 0.05, df_lack_of_fit, df_residuals)

# Calculate p-value for lack of fit
p_value_lack_of_fit <- pf(F_lack_of_fit, df_lack_of_fit, df_residuals, lower.tail = FALSE)

# Print results
cat("F-critical for LOF:", F_critical, "\n")
cat("F-value for lack of fit:", F_lack_of_fit, "\n")
```


# Cross Validation
```{r}
library(caret)

# Leave-One-Out-Cross-Validation
train_control_loocv <- trainControl(method = "LOOCV")
loocv <- train(best_model_formula, data = climate, method = "lm", trControl = train_control_loocv)

# Calculate MSE for LOOCV
mse1 <- loocv$results$RMSE^2
cat("LOOCV MSE:", mse1, "\n")

# K-fold cross-validation
train_control_kfold <- trainControl(method = "cv", number = 10)
kfold <- train(best_model_formula, data = climate, method = "lm", trControl = train_control_kfold)

# Calculate MSE for K-fold CV
mse2 <- kfold$results$RMSE^2
cat("K-fold MSE:", mse2, "\n")
```


# Additional EXPLORATORY Analysis
```{r}
split_month <- climate %>%
  split(.$Month) %>%
  lapply(function(.) {
    .[order(.$Year, decreasing = FALSE), ]
  })

# split_month
### NOT PRINTED AS THE OUTPUT IS TOO LONG
```

```{r}
aggregate_temperature <- function(df) {
  
  # Create a new column that groups years into sets of 5
  df <- df %>% 
    mutate(YearGroup = (row_number() - 1) %/% 5 + 1)
  
  # Calculate the average temperature for each group and rename the year group
  result <- df %>% 
    group_by(YearGroup) %>%
    summarise(
      YearRange = paste(min(Year), max(Year), sep = " - "), 
      Temperature = mean(Temp),
      Month = first(Month) # unchanged
    ) %>%
    ungroup()
  
  return(result)
}

# Apply the function to each split_month dataframe
bymonth <- lapply(split_month, aggregate_temperature)
# bymonth
### NOT PRINTED AS THE OUTPUT IS TOO LONG
```

# FIGURE 1 Exploration of How the Categorical Variable of Month Affects Temperature by Year
```{r}
by_yeargroup <- list()

for (i in 1:5) {
  new_df <- do.call(rbind, lapply(bymonth, function(df) df[i, ]))
  by_yeargroup[[i]] <- new_df
}
# by_yeargroup
### NOT PRINTED AS THE OUTPUT IS TOO LONG

plot <- ggplot() +
  labs(title = "Temperature by Month According to Year Range",
       x = "Month",
       y = "Temperature") +
  scale_color_discrete(name = "Year Range") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +  # Set x-axis breaks and labels
  theme_minimal()

# Iterate through each data frame in the by_yeargroup list and add a geom_line() for each
for (i in 1:length(by_yeargroup)) {
  plot <- plot + geom_line(data = by_yeargroup[[i]], aes(x = Month, 
                                      y = Temperature, color = factor(YearRange)))
}

plot
```




